---
title: "Statistical Analysis of US Border Crossings Data"
output: html_document
geometry: margin=0in
---

#### Importing all the required libraries
```{r}

library(dplyr)
library(tidyverse)
library(ggplot2)
library(stringr)
library(lubridate)
library(gridExtra)
library(scales)
library(fitdistrplus)

```

#### Reading the dataset
```{r}
border_crossings<-read.csv("Border_Crossing_Entry_Data.csv", header=TRUE, sep=',')
head(border_crossings)

```

#### Selecting only the relevant columns
```{r}

filtered_data<-subset(border_crossings,select=-Port.Code)
head(filtered_data)

```

#### Creating separate columns for Month & Year for further processing
```{r}
filtered_data$Month<-day(filtered_data$Date)
filtered_data$Year<-year(filtered_data$Date)

head(filtered_data)

```

#### Finding out the spread across year for our dataset - 1996-2020
```{r}

years<-unique(filtered_data$Year)
years

months<-unique(filtered_data$Month)
months

```

#### Grouping based on Border and Measure to find the individual Frequency count of Inbound Crossings
```{r}

total_value<-filtered_data%>%
             group_by(Border,Measure)%>%
             summarise(Frequency=sum(Value))%>%
             group_by(Border)%>%
             summarise(Total=sum(Frequency))%>%
             arrange(desc(Total))
total_value

```

### Visualizing the total frequency count of inbound crossings at both the borders
```{r warning=FALSE}

options(scipen=999)

ggplot(data=total_value,aes(x=reorder(Border,-Total),y=Total,fill=Border))+geom_bar(position="dodge",stat = "identity") + scale_fill_brewer(palette="Dark2") + ggtitle("Frequency of Inbound Crossings between 1996 and 2020") + ylab("Frequency")+xlab("Border")+theme(plot.title = element_text(size = 12,face="bold"))+
  theme(plot.title = element_text(hjust = 0.5))+ theme(axis.text.x = element_text(size=11,face="bold"))+ theme(axis.text.y = element_text(size=11,face="bold"))
 
```


#### Checking the frequency count of each Measure at both the borders
```{r}

total_value_measure<-filtered_data%>%
             
  group_by(Border,Measure)%>%
             summarise(Frequency=sum(Value))%>%
             arrange(desc(Frequency))
total_value_measure

```

#### Segregating data for personal modes of transportation 
```{r warning=FALSE}

total_measures_general<-filtered_data%>%
             dplyr::filter(Measure %in% c("Personal Vehicle Passengers","Personal Vehicles","Pedestrians"))%>%
             group_by(Border,Measure)%>%
             summarise(Frequency=sum(Value))%>%
             arrange(desc(Frequency))

total_measures_general

```

```{r}

ggplot(data=total_measures_general,aes(y=Frequency,x=reorder(Measure,-Frequency),fill=Border))+geom_bar(stat="identity",position="dodge",color="black")+ggtitle("Frequency of Inbound Crossings by Measure for Personal Modes of Transportation")+ylab("Frequency") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size=11))+ theme(axis.text.y = element_text(size=11))+theme(plot.title = element_text(size = 11,face="bold"))+
  theme(plot.title = element_text(hjust = 0.5))+xlab("Measure")+scale_fill_brewer(palette="Dark2")+ theme(axis.text.x = element_text(size=11,face="bold"))+ theme(axis.text.y = element_text(size=11,face="bold"))

```

#### Segregating data for commerical modes of transportation 
```{r}

total_measures_commercial<-filtered_data%>%
             dplyr::filter(!Measure %in% c("Personal Vehicle Passengers","Personal Vehicles","Pedestrians"))%>%
             group_by(Border,Measure)%>%
             summarise(Frequency=sum(Value))%>%
             arrange(desc(Frequency))

ggplot(data=total_measures_commercial,aes(y=Frequency,x=reorder(Measure,-Frequency),fill=Border))+geom_bar(stat="identity",position="dodge",color="black")+ggtitle("Frequency of Inbound Crossings by Measure for Commercial Modes of Transportation")+ylab("Frequency") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+theme(plot.title = element_text(size = 11,face="bold"))+
  theme(plot.title = element_text(hjust = 0.5))+xlab("Measure")+scale_fill_brewer(palette="Dark2")+ theme(axis.text.x = element_text(size=11,face="bold"))+ theme(axis.text.y = element_text(size=11,face="bold"))

```

#### Visual analysis of Yearly trends
```{r warning=FALSE}

total_value_measure_year<-filtered_data%>%
             
  group_by(Border,Year)%>%
             summarise(Frequency=sum(Value))%>%
             arrange(desc(Frequency))
total_value_measure_year


ggplot(data=total_value_measure_year,aes(y=Frequency,x=Year))+geom_line(aes(color=Border))+geom_point(aes(color=Border))+ylab("Frequency") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+theme(plot.title = element_text(size = 12,face="bold"))+
  theme(plot.title = element_text(hjust = 0.5))+ scale_x_discrete(limits=c(total_value_measure_year$Year))+scale_color_brewer(palette="Dark2")+ggtitle("Frequency of Inbound Crossings by Border & Year")+ theme(axis.text.x = element_text(size=11,face="bold"))+ theme(axis.text.y = element_text(size=11,face="bold"))

```

#### Visual analysis of Monthly trends
```{r}
total_value_measure_month<-filtered_data%>%
             
  group_by(Border,Month)%>%
             summarise(Frequency=sum(Value))%>%
             arrange(desc(Frequency))
total_value_measure_month

```

#### Analyzing Frequency of Inbound Crossings by Month & Border - the highest number of inbound crossings at US-Mexico Border were in the month of December & at the US-Canada Border were in the month of August.
```{r}

ggplot(data=total_value_measure_month,aes(y=Frequency,x=Month,fill=Border))+geom_bar(stat="identity",position="dodge",color="black")+ggtitle("Frequency of Inbound Crossings by Month & Border")+ylab("Frequency") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+theme(plot.title = element_text(size = 12,face="bold"))+
  theme(plot.title = element_text(hjust = 0.5))+ scale_x_discrete(limits=c(total_value_measure_month$Month))+scale_fill_brewer(palette="Dark2")+ theme(axis.text.x = element_text(size=11,face="bold"))+ theme(axis.text.y = element_text(size=11,face="bold"))


```

#### Analyzing frequency of Inbound Crossings by State & Border
```{r}

total_value_state<-filtered_data%>%
             group_by(Border,State)%>%
             summarise(Inbound_Crossings=sum(Value))%>%
             arrange(desc(Inbound_Crossings))

total_value_state

```

```{r}

ggplot(data=total_value_state,aes(y=Inbound_Crossings,x=reorder(State,-Inbound_Crossings),fill=Border))+geom_bar(stat="identity",color="black")+ggtitle("Frequency of Inbound Crossings by State & Border")+ylab("Frequency") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+theme(plot.title = element_text(size = 12,face="bold"))+theme(plot.title = element_text(hjust = 0.5))+ scale_x_discrete(limits=c(total_value_state$State))+xlab("State")+scale_fill_brewer(palette="Dark2")+ theme(axis.text.x = element_text(size=11,face="bold"))+ theme(axis.text.y = element_text(size=11,face="bold"))

```

#### Conditional Probability

1. What is the probability that a truck entering the US will go to El Paso port given that it is crossing the US-Mexico border?

A = event that a truck has entered visa the US-Mexico Border
B = event that a truck goes to El Paso

P(𝐵|𝐴) = 𝑃(𝐴∩𝐵) /𝑃(𝐴)

```{r}
#unique(Mexico$Port.Name)

# Truck entering via US-Mexico Border goes to EL Paso Port
El_Paso<-subset(filtered_data,(Border=="US-Mexico Border" & Port.Name=="El Paso"& Measure=="Trucks"))
total_el_paso<-El_Paso%>%group_by(Port.Name)%>%summarise(Total=sum(Value))

# Total trucks entering the US
Trucks<-subset(filtered_data, Measure=="Trucks")
total_trucks<-Trucks%>%group_by(Measure)%>%summarise(Total=sum(Value))

#Probability that a truck entering the US will go to El Paso port given that it is crossing the US-Mexico border
total_el_paso$Total/total_trucks$Total

```

2. What is the probability that a person that has entered the US for a non-business trip via US-Canada Border given that the person is a Pedestrian?

A = event that a person has entered the US on a non-business trip(Bus, Pedestrian, etc.)
B = event that a person is a Pedestrian amongst all non-business persons

P(𝐵|𝐴) = 𝑃(𝐴∩𝐵) /𝑃(𝐴)

```{r}
# Total pedestrians entering the US via US-Canada Border
Pedestrians<-subset(filtered_data, Border=="US-Canada Border" & Measure=="Pedestrians")
total_pedestrians<-Pedestrians%>%group_by(Measure)%>%summarise(Total=sum(Value))

# Total Pedestrians entering the US regardless of Border
Total_Pedestrians<-subset(filtered_data, Measure=="Pedestrians")
total_pedestrians_us<-Total_Pedestrians%>%group_by(Measure)%>%summarise(Total=sum(Value))

#probability that a person entering the US via US-Canada Border is a Pedestrian
total_pedestrians$Total/total_pedestrians_us$Total

```


#### Fitting Distribution - Visualizing the distribution 
```{r}
canada_dist <-filtered_data %>%
              dplyr::filter(Border=="US-Canada Border") %>%
              group_by(Value) %>%
              summarise(No_of_Inbound_Crossings_Canada = n())
              
canada_dist

ggplot(canada_dist, aes(Value)) +
geom_histogram(bins = 50, color = 'Black', fill = 'brown')+ggtitle("Distribution of Inbound Crossings at US-Canada Border")+ylab("Frequency")+xlab("No of Inbound Crossings")+scale_fill_brewer(palette="Dark2")+ theme(axis.text.x = element_text(size=11,face="bold"))+ theme(axis.text.y = element_text(size=11,face="bold"))


```

```{r}
mexico_dist <-filtered_data %>%
              dplyr::filter(Border=="US-Mexico Border") %>%
              group_by(Value) %>%
              summarise(No_of_Inbound_Crossings_Mexico = n())
              
mexico_dist

ggplot(mexico_dist, aes(Value)) +
geom_histogram(bins = 50, color = 'Black', fill = 'brown')+ggtitle("Distribution of Inbound Crossings at US-Mexico Border")+ylab("Frequency")+xlab("No of Inbound Crossings")+scale_fill_brewer(palette="Dark2")+ theme(axis.text.x = element_text(size=11,face="bold"))+ theme(axis.text.y = element_text(size=11,face="bold"))

```

#### CANADA BORDER - Negative Binomial distibution 
```{r}
fit_nb_can <- fitdist(canada_dist$No_of_Inbound_Crossings_Canada, 'nbinom',discrete = TRUE)
summary(fit_nb_can)
```

#### CANADA BORDER - Poisson distribution
```{r}
fit_p_can<- fitdist(canada_dist$No_of_Inbound_Crossings_Canada, 'pois',discrete = TRUE) 
summary(fit_p_can)
```

#### From the below goodness-of-fit test, and lower AIC and BIC values and higher Loglikelihood,we infer that normal
distribution is a better fit for the Number of Inbound Crossings variable for US-Canada Border with a µ = 8.257.

```{r}

gofstat(list(fit_nb_can, fit_p_can))

```

#### MEXICO BORDER - Negative Binomial Distribution

```{r}

# Getting the parameter estimates for negative binomial distribution
fit_nb_mex <- fitdist(mexico_dist$No_of_Inbound_Crossings_Mexico, 'nbinom',discrete = TRUE)
summary(fit_nb_mex)

```

### MEXICO BORDER - Poisson Distribution

```{r}
# Getting the parameter estimates for Poisson distribution
fit_p_mex <- fitdist(mexico_dist$No_of_Inbound_Crossings_Mexico, 'pois',discrete = TRUE) 
summary(fit_p_mex)

```

#### From the below goodness-of-fit test, and lower AIC and BIC values and higher Loglikelihood,we infer that normal
distribution is a better fit for the Number of Inbound Crossings variable for US-Mexico Border with a µ = 2.567.

```{r}

# Implementing the goodness-of-fit tests
gofstat(list(fit_nb_mex, fit_p_mex))

```

### STATISTICAL ANALYSIS

### 1. One sample Z-test : Assuming the entire dataset is the population, we want to test if the mean value of inbound crossings of our sample data is equivalent to the population mean value of inbound crossings. 

#### Population Mean
```{r}
cat("Population Mean =" ,mean(filtered_data$Value))

```

X ≡ R.V. of number of inbound crossings in the U.S.
Null Hypothesis ->      H0: µ = 28302.55 
Alternate Hypothesis -> H1: µ != 28302.55

#### Performing the One-Sample Z-test
```{r}

# Step 1: Create a z-test function manually.

##> This function has two input arguments (sample, pop)
##> The output of this function is the calculated Z_calc and P_value)

# Create the function: one sample z test
z.test <- function(sample, pop){
  sample_mean = mean(sample) 
  pop_mean = mean(pop)
  n = length(sample) 
  var = var(pop)
  z = (sample_mean - pop_mean) / sqrt((var/(n))) 
  df<-data.frame("Z_calc"=z,"P_value"=pnorm(z))
  return(df)
}


#Step 2: Create samples

#to return the same data each time we sample it, we need to set the seed to a numerical value
set.seed(100) 

#sample data
borderSample <- sample_n(filtered_data, 1000)


sample     = borderSample$Value
population = filtered_data$Value

cat("Sample Mean = ", mean(sample))

# Step 3: Run the test 
cat(sep="\n")
z.test(sample, population)

```
#### Conclusion of One sample Z-test : As P_value < 0.05 we reject the null hypothesis and thus conclude there
is a significant difference between the sample mean value of inbound crossings and population mean value of inbound crossings.





### 2. Two sample Z-test : To test if the mean value of inbound crossings of two different set of samples taken from two different populations is the same or not

X1 ≡ R.V. of inbound crossings in the U.S. from first sample
X2 ≡ R.V. of inbound crossings in the U.S. from second sample

Null Hypothesis ->      H0: µ1 − µ2 = 0
Alternate Hypothesis -> H1: µ1 − µ2 != 0

```{r}

# Step 1: Create a two sample z-test function manually.

z_test2 = function(a, b, var_a, var_b){
  n.a = length(a)
  n.b = length(b)
  z = (mean(a) - mean(b)) / (sqrt((var_a)/n.a + (var_b)/n.b))
  df<-data.frame("Zcal"=z,"P_value"=pnorm(z))
  return(df)
}

#Step 2: Create two samples

#to return the same data each time we sample it, we need to set the seed to a numerical value
set.seed(100)
# 
border_1 <- filtered_data[1:181220,] #select all columns and rows from 1 to 181220  
border_2 <- filtered_data[181221:362440,] #select all columns and rows from 181221 to 362440

border_1_sample <- sample_n(border_1, 10000) #sample 10000 rows from the first population
border_2_sample <- sample_n(border_2, 10000) #sample 10000 rows from the second population

sample1<-border_1_sample$Value
# The mean value of inbound crossings of sample 1 is (µ1) = 23701.77
cat("Sample Mean 1 = ",mean(sample1))

cat(sep="\n")
sample2<-border_2_sample$Value
# The mean value of inbound crossings of sample 2 is (µ2) = 32351.66
cat("Sample Mean 2 = ",mean(sample2))



var_1<-var(border_1_sample$Value)
var_2<-var(border_2_sample$Value)

# Step 3: Run the test

cat(sep="\n")
z_test2(sample1,sample2,var_1,var_2)

```
#### Conclusion of One sample Z-test : As P_value < 0.05 we reject the null hypothesis and thus conclude there
is a significant difference between the sample mean value of inbound crossings of the two samples. 



### 3. Two sample t-test : To test the significance between means of two samples means if populations
are independent and the variances are unknown, i.e., to test the equality of the mean value of inbound crossings of two random samples generated from the border crossing data.

X1 ≡ R.V. of inbound crossings in the U.S. from first sample
X2 ≡ R.V. of inbound crossings in the U.S. from second sample

Null Hypothesis ->      H0: µ1 − µ2 != 0
Alternate Hypothesis -> H1: µ1 − µ2 = 0

```{r}

#Step 1: Create two samples

#to return the same data each time we sample it, we need to set the seed to a numerical value
set.seed(100)

border_1 <- filtered_data[1:181220,] #select all columns and rows from 1 to 181220  
border_2 <- filtered_data[181221:362440,] #select all columns and rows from 181221 to 362440
border_1_sample <- sample_n(border_1, 10000) #sample 10000 rows from the first population
border_2_sample <- sample_n(border_2, 10000) #sample 10000 rows from the second population

sample1<-border_1_sample$Value
cat("Sample Mean 1 = ",mean(sample1))

cat(sep="\n")
sample2<-border_2_sample$Value
cat("Sample Mean 2 = ",mean(sample2))

#Step - 2 : performing the t test

cat(sep="\n")
t.test(x=sample1,y=sample2)


```

#### Conclusion of Two sample t-test : As P_value > 0.05 we fail to reject the null hypothesis and thus conclude there
is no significant difference between the sample mean value of inbound crossings of the two samples. 


### 4. Two sample proportion test : To test the equality of proportions of inbound bus crossings in New York at US-Canada Border & in Texas at US-Mexico Border in 2020

```{r}

mexico_data<-filtered_data%>%
             dplyr::filter(Border=="US-Mexico Border")%>%
             group_by(Border,State,Measure,Year)%>%
             summarise(Inbound_Crossings=sum(Value))
mexico_data

canada_data<-filtered_data%>%
             dplyr::filter(Border=="US-Canada Border")%>%
             group_by(Border,State,Measure,Year)%>%
             summarise(Inbound_Crossings=sum(Value))
 
canada_data

```


X1 ≡ R.V. of number of inbound bus crossings at US-Canada Border
X2 ≡ R.V. of number of inbound bus crossings at US-Mexico Border

Null Hypothesis ->      H0: p1 − p2 = 0 
Alternate Hypothesis -> H1: p1 - p2!= 0

```{r warning=FALSE}

# Proportion inbound buses at US-Canada ==inbound buses at US-Mexico Border

# Step 1-a: Determine the number of inbound buses at US-Canada Border
n1 <-length(which(canada_data$Measure == "Buses"))
cat("Number of inbound buses at the US-Canada Border(n1) = ",n1)
cat(sep="\n")

# Step 1-b: Filter and calculate the number of inbound buses at US-Canada Border in the State of New York in 2020
x1 <- canada_data %>% dplyr:: filter(Measure == "Buses" , State=="New York") %>% nrow()
cat("Number of inbound buses at the US-Canada Border in the State of New York in 2020(x1) = ",x1)
cat(sep="\n")

# Step 1-c: Determine the number of inbound buses at US-Mexico Border
n2 <-length(which(mexico_data$Measure == "Buses"))
cat("Number of inbound buses at the US-Mexico Border(n2) = ",n2)
cat(sep="\n")

# Step 1-d: Filter and calculate the number number of inbound buses at US-Mexico Border in the State of Texas in 2020  
x2 <- mexico_data %>% dplyr:: filter(Measure == "Buses" , State=="Texas") %>% nrow()
cat("Number of inbound buses at the US-Mexico Border in the State of Texas in 2020(x2) = ",x2)
cat(sep="\n")

# step 2: Execute the test
prop.test(x=c(x1, x2),n=c(n1,n2))

```

#### Conclusion of Two sample proportion test : As P_value < 0.05 we reject the null hypothesis and thus conclude there
is a significant difference between the proportions of inbound bus crossings in New York at US-Canada Border & in Texas at US-Mexico Border in 2020



